Linear response theory for a system under non-equilibrium conditions

### 2013-06-28 writedown/linear.response/paper.tex
changed to Mol. Phys. template

% \documentclass[aip,jcp,preprint,unsortedaddress,a4paper,onecolum]{revtex4-1}
\documentclass[aip,jcp,a4paper,reprint,onecolumn]{revtex4-1}
% \documentclass[aps,pre,twocolumn]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

\usepackage[fleqn]{amsmath}
\usepackage{amssymb,amsthm}
\usepackage[dvips]{graphicx}
\usepackage{color}
\usepackage{tabularx}
\usepackage{algorithm}
\usepackage{algorithmic}

@@

% \title{Linear response theory for core set identification}
\title{Linear response theory for a system under non-equilibrium conditions}
\author{Han Wang}
\author{Carsten Hartmann}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\author{Christof Sch\"utte}
\affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
\affiliation{Zuse Institute Berlin (ZIB), Germany}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

\date{\today}

@@

Here we propose an alternative (purely formal) derivation of the linear response result, that is based on a change of drift in the corresponding SDE and which allows for an easy generalization of the above linear response result to nonequilibrium systems. We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook \cite{oksendal}. Now let $x_{t},X_{t}$ be the solutions to the following stochastic differential equations:
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t} \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t} \label{sde2-2}
\end{align}
\end{subequations}
for $0\le t\le T$ and with fixed initial conditions
\[
x_{0} = X_{0} = x\,,
\]
Suppose that there exists a auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,,
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to express (\ref{sde2-2}) by
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any function $f=f(\{X_{t}\}_{0\le t\le T})$ that is integrable \wrt $\nu$, we have the identity
\[
\bE_{\nu}[f] := \int f(\{X_{t}\}_{0\le t\le T})\, d\nu = \int f(\{X_{t}\}_{0\le t\le T}) \frac{d\nu}{d\mu}\,d\mu =: \bE_{\mu}[M^{-1}_{T}f]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)\,.
\]
is the density of $\nu$ relative to $\mu$.

@@

% \newpage
% \section{Numerical results}
% \subsection{Tilting double-well potential}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.4\textwidth]{figs/fig-tilt-pot.eps}
%   \includegraphics[width=0.4\textwidth]{figs/fig-split-pot.eps}
%   \caption{Two testing cases presented by this paper.
%     Left: the double-well potential with tilting perturbation.
%     Right: the single-well potential with splitting perturbation.}
%   \label{fig:new-tmp1}
% \end{figure}

% We test the idea by a one-dimensional model system: one particle in a
% tilting double-well potential. For convenience, we let the mass of the
% particle to be 1 \textsf{amu}. The unperturbed
% Hamiltonian of the system is given by:
% \begin{align}
%   \mh ( p,  q) = \frac 12  p^2 + U( q) 
% \end{align}
% with potential
% \begin{align}
%   U( q) = \frac12 k ( q^2 - a^2)^2
% \end{align}
% Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^4)$, and $ a = 1\ \textsf{nm}$.
% Notice at room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
% See the red line in Fig.~\ref{fig:tmp1}.
% %$\textsf{kJ} / (\textsf{mol nm}^4)}$
% The perturbation is given by
% \begin{align}
%    C( p,  q) = 0; \qquad
%    D( p,  q) = -\nabla_{ q} V( q) = 1
% \end{align}
% Here $V( q) = - q$ is  effectively tilting the original
% potential $U( q)$. The strength of the perturbation $F_e(t)$ is given
% by the following function:
% \begin{align}
%   F_e(t) = 
%   \begin{cases}
%     F_e^{\textrm{max}}\times (t / t_c) & t < t_c \\
%     F_e^{\textrm{max}} & t \geq t_c
%   \end{cases}
% \end{align}
% $t_c = 20$~\textsf{ps} so that the perturbation increases slow
% enough: the typical decaying time scale of the correlation function
% in~\eqref{eqn:core-identify-approx-2} is roughly 3~\textsf{ps}.
% We choose $\tau = 1$~\textsf{ps}.
% % We choose $\mt_\alpha$
% % to be the propagator of the Langevin dynamics at temperature $150$
% % \textsf{K} with $\alpha = 1$ \textsf{ps}.
% Since the dynamics is in
% 1-d, the projection $\mathcal P$ is identity.  We consider the stability of the
% distribution $f( x, t)$ at time $t$: $\Delta\mt_\tau f( x, t)$.
% The larger
% this value, the more stable the core sets are.

% \begin{figure}
%   \centering
%   \includegraphics[width=0.95\textwidth]
%   {figs/fig-tilt-str2-simple.eps}
%   \caption{
%     Tilting double well potential testing case:
%     the plot of $\Delta\mt_\tau f( x,t)$  under perturbation of
%     $F_e^{\textrm{max}} = 2.0$.
%     From left to right:
%     (a) $t=0$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (b) $t=20$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (c) $t=20$~\textsf{ps},
%     response formula truncated at order $n=1$,
%     and
%     (d) $t=20$~\textsf{ps},
%     response formula truncated at order $n=2$.
%     The equilibirum state serves as the reference of
%     the response simulations.
%   }
%   \label{fig:new-tmp2}
% \end{figure}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.95\textwidth]
%   {figs/fig-tilt-str4-simple.eps}
%   \caption{
%     Tilting double well potential:
%     the plot of $\Delta\mt_\tau f( x,t)$  under perturbation of
%     $F_e^{\textrm{max}} = 4.0$.
%     From left to right:
%     (a) $t=0$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (b) $t=20$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (c) $t=20$~\textsf{ps},
%     response formula with reference process $F_e^{\max} = 0.0$
%     (i.e. equilibirum simulation),
%     and
%     (d) $t=20$~\textsf{ps},
%     response formula with reference process $F_e^{\max} = 2.0$.
%     The response formula are truncated at order $n=1$.
%   }
%   \label{fig:new-tmp3}
% \end{figure}

% The Fig.~\ref{fig:new-tmp2} and \ref{fig:new-tmp3} present
% the application of the response theory to
% calculate core sets of the tilting double-well potential.
% Fig.~\ref{fig:new-tmp2} presents
% the snapshot at $t=20$~\textsf{ps}
% of $\Delta\mt_\tau f( x,t)$ under the maximum perturbation
% strength $F_e^{\max} = 2.0$. Response formula truncated
% at both order $n=1$ and $n=2$ are compared to 
% the brutal force non-equilibirum simulation.
% No obviously difference is shown. The second order response result
% is slightly more precise than the first order one, but the
% statistical error is also larger, which makes the profile noisy.
% This figure tells us that
% (1) the modeling error of the linear response (1st order) is
% negligibly small; (2) the statistical error of numerically measuring the
% linear response is also small.
% Fig.~\ref{fig:new-tmp3} presents the $t=20$~\textsf{ps}
% snapshots of $\Delta\mt_\tau f( x,t)$
% under a larger perturbations $F_e^{\max} = 4.0$.
% The linear reponse using the equilibirum process as the reference
% shows poor accuracy: the left well vanishes in this case,
% while the linear response produces  artifical structures.
% Since the profile is smooth, which indicates
% the statistical error of calculating the linear term is small,
% the deviation comes from the modellng error, namely the contribution
% of higher order response terms.
% By including the second order response term reduces the left
% artifical core set, but also creates artifical finer grained
% structures. The statistical accuracy is also worse than the linear response.
% Plot (d) present the result of the linear response
% formula using reference process $F_e^{\max} = 2.0$, which
% is the process presented in Fig.~\ref{fig:new-tmp2}.
% Much better modelling accuracy is achieved:
% the left artifical core set disappear and the
% statistical accuracy is also satisfactory.

% % See Fig. \ref{fig:tmp2} and \ref{fig:tmp3} for numerical result.  Both
% % the direct non-equilibrium simulation and the response
% % approximations are presented. Time slices $t = 0$, 10, 15 and 20
% % \textsf{ps} are shown.  When the perturbation is small,
% % i.e. $F_e^{\textrm{max}} = 2$, the direct non-equilibrium simulation
% % and the linear response results are accurately consistent.
% % The 2nd order response formula~\eqref{eqn:pert-approx-1} and
% % \eqref{eqn:pert-approx-2} are even better, but the numerical error
% % is bigger.
% % As the
% % perturbation grows stronger, the left well becomes less stable, while
% % the right well becomes more stable.  When the perturbation is big,
% % i.e. $F_e^{\textrm{max}} = 4$, the linear response result is no
% % longer precise.
% % When $t\geq 20$ \textsf{ps}, the left well basically
% % disappears, however, the linear response calculation presents
% % artifical occilation at the right well.
% % The 2nd order response also has artifical effect, and the numerical
% % uncertainty is high.
% % The 4th rwo of Fig.~\ref{fig:tmp3} presents the 1st order result of the
% % response formula~\eqref{eqn:pert-approx-1} and
% % \eqref{eqn:pert-approx-2}, with
% % respect to the reference state $F_e^{\textrm{max}} = 2$. The result
% % is impressively improved compared with the linear response with respect to the
% % equilibrium state.

% \subsection{Splitting single-well potential}

% In this splitting single-well potential example, the unperturbed
% potential is simply the harmonic potential given by:
% \begin{align}
%   U( q) = \frac12\,k\, q^2 
% \end{align}
% Here $k = 8$~$\textsf{kJ} / (\textsf{mol nm}^2)$.
% Notice at room temperature $300\ \textsf{K}$, $k_BT = 2.48$~\textsf{kJ/mol}.
% See the red line in Fig.~\ref{fig:tmp4}.
% %$\textsf{kJ} / (\textsf{mol nm}^4)}$
% The perturbation is again given by
% \begin{align}
%    C( p,  q) = 0; \qquad
%    D( p,  q) = -\nabla_{ q} V( q) 
% \end{align}
% where $V( q)$ is a Gaussian type potential:
% \begin{align}
%   V( q) = \frac{1}{\sqrt{2\pi \sigma^2}}
%   \exp\Big\{-\frac{ q^2}{2\sigma^2}\Big\}
% \end{align}
% we use $\sigma = 0.16$~\textsf{nm}. Please see Fig~\ref{fig:tmp4}.
% For the results, please see Fig.~\ref{fig:tmp5} -- Fig.~\ref{fig:tmp6}.

% \begin{figure}
%   \centering
%   \includegraphics[width=0.95\textwidth]
%   {figs/fig-split-str1-simple.eps}
%   \caption{
%     Splitting single-well potential testing case:
%     the plot of $\Delta\mt_\tau f( x,t)$  under perturbation of
%     $F_e^{\textrm{max}} = 1.0$.
%     From left to right:
%     (a) $t=0$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (b) $t=20$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (c) $t=20$~\textsf{ps},
%     response formula truncated at order $n=1$,
%     and
%     (d) $t=20$~\textsf{ps},
%     response formula truncated at order $n=2$.
%     The equilibirum state serves as the reference of
%     the response simulations.
%   }
%   \label{fig:new-tmp5}
% \end{figure}

% \begin{figure}
%   \centering
%   \includegraphics[width=0.95\textwidth]
%   {figs/fig-split-str2-simple.eps}
%   \caption{
%     splitting single well potential:
%     the plot of $\Delta\mt_\tau f( x,t)$  under perturbation of
%     $F_e^{\textrm{max}} = 2.0$.
%     From left to right:
%     (a) $t=0$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (b) $t=20$~\textsf{ps},
%     the brutal force non-equilibirum simulation,
%     (c) $t=20$~\textsf{ps},
%     response formula with reference process $F_e^{\max} = 0.0$
%     (i.e. equilibirum simulation),
%     and
%     (d) $t=20$~\textsf{ps},
%     response formula with reference process $F_e^{\max} = 1.0$.
%     The response formula are truncated at order $n=1$.
%   }
%   \label{fig:new-tmp6}
% \end{figure}

\section*{References}
\bibliography{ref}{}
\bibliographystyle{unsrt}

>>>>

% \documentclass[aip,jcp,preprint,unsortedaddress,a4paper,onecolum]{revtex4-1}
\documentclass[]{tMPH2e}
% \documentclass[aps,pre,twocolumn]{revtex4-1}
% \documentclass[aps,jcp,groupedaddress,twocolumn,unsortedaddress]{revtex4}

% \usepackage[fleqn]{amsmath}
% \usepackage{amssymb,amsthm}
% \usepackage[dvips]{graphicx}
% \usepackage{color}
% \usepackage{tabularx}
% \usepackage{algorithm}
% \usepackage{algorithmic}

@@

% \title{Linear response theory for core set identification}
\title{Linear response theory for a system under non-equilibrium conditions}

\author{Han Wang$^{a}$ and Carsten Hartmann$^{a}$ and Christof Sch\"utte$^{a,b}$$^{\ast}$\thanks{$^\ast$Corresponding author. Email: Christof.Schuette@fu-berlin.de}\\\vspace{6pt}$^{a}${\em Institute for Mathematics, Freie Universit\"at Berlin, Germany}\\$^{b}${\em Zuse Institute Berlin (ZIB), Germany}}

% \author{Han Wang}
% \author{Carsten Hartmann}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
% \author{Christof Sch\"utte}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}
% \affiliation{Zuse Institute Berlin (ZIB), Germany}
% \affiliation{Institute for Mathematics, Freie Universit\"at Berlin, Germany}

@@

Here we propose an alternative (purely formal) derivation of the linear response result, that is based on a change of drift in the corresponding SDE and which allows for an easy generalization of the above linear response result to nonequilibrium systems. We will briefly review the idea of the change of drift via Girsanov transformations; for details we refer to the textbook~\cite{oksendal2003stochastic}. Now let $x_{t},X_{t}$ be the solutions to the following stochastic differential equations:
\begin{subequations}\label{sde2}
\begin{align}
dx_{t} & = b(x_{t},t)dt + a(x_{t})dB_{t} \label{sde2-1}\\
dX_{t} & = (b(X_{t},t) + \eps v_{t})dt +a(X_{t})dB_{t} \label{sde2-2}
\end{align}
\end{subequations}
for $0\le t\le T$ and with fixed initial conditions
\[
x_{0} = X_{0} = x\,,
\]
Suppose that there exists a auxiliary stochastic process $\xi_{t}\in\R^{m}$ such that 
\begin{equation}\label{xi}
a(X_{t})\xi_{t} = v_{t}\,,
\end{equation}
The auxiliary variable $\xi$ will be called \emph{control variable}. We define  
\[
W_{t} = \eps\int_{0}^{t}\xi_{s}\,ds + B_{t}\,,\quad 0\le t\le T\,,
\]
which allows us to express (\ref{sde2-2}) by
\begin{equation}\label{sde2-3}
dX_{t} = b(X_{t},t)dt +a(X_{t})dW_{t} 
\end{equation}
It follows from the Girsanov theorem \cite[Thm.~8.6.8]{oksendal2003}, sometimes also called \emph{Cameron-Martin-Girsanov theorem} \cite{a}, that $W_{t}$ is again a Brownian motion under a new probability measure that has a density \wrt the Gaussian probability measure that is generated by the Brownian motion $B_{t}$. Specifically, let $\nu$ denote the law of the Brownian motion $B_{t}$ and define a new probability measure $\mu$ on the space of continuous trajectories by 
\[
d\mu=M_{T}d\nu
\]
with 
\begin{equation}\label{likelihood}
M_{t} = \exp\left(-\eps\int_{0}^{t}\xi_{s}\cdot dB_{s} - \frac{\eps^{2}}{2}\int_{0}^{t}|\xi_{s}|^{2}ds\right)\,,\quad 0\le t\le T\,.
\end{equation}
Technical details aside, the Girsanov theorem implies that $W_{t}$  for any function $f=f(\{X_{t}\}_{0\le t\le T})$ that is integrable \wrt $\nu$, we have the identity
\[
\bE_{\nu}[f] := \int f(\{X_{t}\}_{0\le t\le T})\, d\nu = \int f(\{X_{t}\}_{0\le t\le T}) \frac{d\nu}{d\mu}\,d\mu =: \bE_{\mu}[M^{-1}_{T}f]\,.
\]
where 
\[
M_{T}^{-1} = \exp\left(\eps\int_{0}^{T}\xi_{s}\cdot dW_{s} - \frac{\eps^{2}}{2}\int_{0}^{T}|\xi_{s}|^{2}ds\right)\,.
\]
is the density of $\nu$ relative to $\mu$.

@@

\bibliography{ref}{}
\bibliographystyle{tMPH}
